A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance and physical activity patterns using unsupervised clustering

### 2019-09-05 draft.tex
more additions to ACT appendix

We also assessed the clustering stability on different data partitions. Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.

\begin{figure}[!htb]
  \centering
  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
  \includegraphics[]{ACTclusteringOn80.png}
  \label{fig:ACTclusteringon80}
\end{figure}

*Not sure about this part. Maybe move to appendix?
We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters. Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count. It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts. This validates that there was a change over time in the patient’s ACT pattern that could be captured. The additional sdBreathAmplitude feature make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.

\begin{figure}[!htb]
  \centering
  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
  \includegraphics[]{ACTclusteringAB.png}
  \label{fig:ACTclusteringAB}
\end{figure}

\subsection{Clustering physical activity waveforms}

@@

\begin{table}[H]
  \caption{Clusters Metrics Kmeans vs Number of Clusters}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Number of Clusters & Average Silhouette width & Calinski-Harabasz index \\
    \midrule
    3 & 2292 & 0.17 \\
    4 & 2208 & 0.18 \\
    5 & 1996 & 0.15 \\
    6 & 1830 & 0.15 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

>>>>

We assessed the clustering stability on different data partitions. Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.

\begin{figure}[!htb]
  \centering
  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
  \includegraphics[]{ACTclusteringOn80.png}
  \label{fig:ACTclusteringon80}
\end{figure}
We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters (see Appendix).
\subsection{Clustering physical activity waveforms}

@@

\begin{table}[H]
  \caption{Clusters Metrics Kmeans vs Number of Clusters}
  \label{steps_metrics}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
    \midrule
    3 & 2292 & 0.17 \\
    4 & 2208 & 0.18 \\
    5 & 1996 & 0.15 \\
    6 & 1830 & 0.15 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\subsection{Hierarchical Clustering}
Figure \ref{fig:figACTdent} shows the Dendrogram of ACT data demonstrating major splits for number of clusterts k = 2,3,5 and 6

\begin{figure}[htb]
  \centering
  \caption{ Hierarchical Clustering Dendrogram for ACT data}
  \includegraphics[scale=0.5]{fig_ACT_HC_dentogram.png}
  \label{fig:figACTdent}
\end{figure}

\subsubsection{Hierarchical Clustering Metrics and Number of Clusters}
The table \ref{tab:act_hc_clusters} shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters.

\begin{table}[htb]
  \caption{Clusters Metrics for Hierahical Clustering vs Number of Clusters}
  \label{tab:act_hc_clusters}
  \centering
  \begin{tabular}{ c|c|c}
    \toprule
    \cmidrule(r){1-2}
    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
    \midrule
    3 & 1793 &  0.13\\
    4 & 1643 &  0.13\\
    5 & 1414 &  0.08\\
    6 & 1289 &  0.06 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{figure}[htb]
  \centering
  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Hierarchical clustering}
  \includegraphics[scale=0.5]{fig_ACT_HC_4_Clusters.png}
  \label{fig:figHC4clusters}
\end{figure}

\section{Appendix : ACT Cluster validations via patient trajectories}
We assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters. Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count. It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts. This validates that there was a change over time in the patient’s ACT pattern that could be captured. The additional breath amplitude standard deviation feature (sdBreathAmplitude) make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.

\begin{figure}[htb]
  \centering
  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
  \includegraphics[]{ACTclusteringAB.png}
  \label{fig:ACTclusteringAB}
\end{figure}
